Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

@@

Empirical research uses evidence from observation and experimentation to
address real-world challenges. In this context, researchers typically
formulate their research questions as \emph{estimands} or \emph{targets
of inference}, i.e., the specific quantities they seek to determine
\citep{Everitt_et_al_2010}. For instance, researchers might be
interested in answering the following question: ``To what extent do
different teaching methods \((T)\) influence students' ability to
produce high-quality written texts \((Y)\)?'' To investigate this,
researchers could randomly assign students to two groups, each exposed
to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
perform pairwise comparisons, generating a dichotomous outcome
\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
ability. In this scenario, the research question can be rephrased as the
estimand, ``\emph{On average}, is there a difference in the ability to
produce high-quality written texts between the two groups of students?''
and this estimand can be mathematically represented by the random
quantity \(E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]\), where
\(E[\cdot]\) denotes the expected value.

@@

While numerous methods can approximate an estimand, researchers
prioritize estimators with desirable identification properties that
ensure the accuracy of estimates. For instance, the Z-test is an
estimator known for its effectiveness when comparing groups'
proportions, yielding accurate estimates when the underlying assumptions
of the statistic are met \citep{Kanji_2006}. If this is the case, the
Z-test is expressed as a signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i} \: | \: T_{i}=1]\) and \(E[Y_{i} \: | \: T_{i}=2]\),
respectively. The noise, represented by \(\hat{s}_{p}\), is defined as
the unpooled sample variability observed between the two groups.

@@

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers to interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key properties, such as
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from any other variable \(X\)
\citep{Morgan_et_al_2014, Neal_2020}.

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendix-A}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes:
\(\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)\).
Here, \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \: | \: do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \: | \: do(T_{i}=1)\) represents
the potential outcome under intervention \(T_{i}=2\). Note that if a
student is assigned to intervention \(T_{i}=1\), the potential outcome
under \(T_{i}=2\) becomes a counterfactual, as it is no longer observed
nor observable. To address the challenge of unobserved counterfactuals,
the approach extends the ICE to the \emph{average causal effect} (ACE):
\(\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]\),
representing the average difference between observed potential outcomes
and their counterfactual counterparts.

Even when data originates from an observational study, researchers can
identify the ACE from associational estimates using the structural
approach. They achieve this by performing statistical conditioning on a
\emph{sufficient adjustment set} of variables \(X\)
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones, ensuring the ACE
estimator satisfies several key properties, including confounding
elimination. If such a set exists, then \(T\) and \(Y\) are
\emph{d-separated} by \(X\) \citep{Pearl_2009}, meaning researchers can
estimate the ACE from associational random quantities
\citep{Morgan_et_al_2014}. Naturally, the validity of claims about the
effects of \(T\) on \(Y\) now hinges on the assumption that \(X\) serves
as a sufficient adjustment set. However, as
\citet[pp.~150]{Kohler_et_al_2019} observed, drawing conclusions about
the real world from observed data inevitably requires assumptions. This
holds true for both observational and experimental data.

For instance, if researchers are unable to conduct the randomized
experiments described in Section~\ref{sec-appendix-A} and must rely on
observational data, they can still estimate the ACE, provided that a
variable \(X\), such as the socio-economic status of the school, blocks
all non-causal paths between the teaching method \(T\) and the outcome
\(Y\) without opening any new ones. Under these conditions, the ACE can
be estimated from the \emph{conditional average causal effect} (CACE),
defined by \(E[Y_{i} \: | \: T_{i}=t, X]\). Using the CACE, the ACE can
be expressed as \(\tau =\)
\(E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =\)
\(E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]\),
where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}. Notably, the approach extends the ACE for a
continuous variable \(T\) as \(\tau =\)
\(E[Y_{i} \: | \: do(T_{i}=t)] =\)
\(d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt\), ensuring broad
applicability across different causal scenarios
\citep[pp.~45]{Neal_2020}.

@@

\newcommand{\dsep}{\perp\!\!\!\perp}

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that help researchers determine which statistical models can
identify an estimand, assuming the depicted causal structure is correct,
thus enabling valid causal inference \citep{McElreath_2020}. They
achieve this by representing various causal structures, regardless of
complexity, using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Furthermore, SCMs and DAGs depict
causal relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
requiring detailed specifications of functional relationships relating
the variables, their parameters, or their data types
\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows the role of
theoretical models in the inference process.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{3} & := f_{X3}(e_{X3}) \\
  e_{X1} & \perp\!\!\!\perp e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
  e_{X1} & \perp\!\!\!\perp e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
  e_{X1} & \perp\!\!\!\perp e_{X2} \\
  e_{X1} & \perp\!\!\!\perp e_{X3} \\
  e_{X2} & \perp\!\!\!\perp e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
  X_{2} & := f_{X2}(e_{X2}) \\
  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
  e_{X1} & \perp\!\!\!\perp e_{X2} \\
  e_{X1} & \perp\!\!\!\perp e_{X3} \\
  e_{X2} & \perp\!\!\!\perp e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
  X_{3} & := f_{X3}(e_{X3}) \\
  e_{X1} & \perp\!\!\!\perp e_{X2} \\
  e_{X1} & \perp\!\!\!\perp e_{X3} \\
  e_{X2} & \perp\!\!\!\perp e_{X3}
\end{aligned}
\]

@@

A careful examination of the figures highlights the theoretical
assumptions underlying these building blocks. SCM \ref{fig-scm_bb1} and
DAG \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
scenario where variables \(X_{1}\) and \(X_{3}\) are independent or not
causally related. SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}
illustrate two connected nodes, representing a scenario where a
\emph{parent} node \(X_{1}\) exerts a causal influence on a \emph{child}
node \(X_{3}\). In this setup, \(X_{3}\) is considered a
\emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because an edge directly connects them.
SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a \emph{chain},
where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences
\(X_{3}\). In this configuration, \(X_{1}\) is a parent node of
\(X_{2}\), which is a parent node of \(X_{3}\). This structure creates a
\emph{directed path} between \(X_{1}\) and \(X_{3}\). Consequently,
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and \(X_{2}\) fully
\emph{mediates} the relationship between the two. SCM \ref{fig-scm_bb4}
and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). Here,
\(X_{2}\) is a parent node that \emph{confounds} the relationship
between \(X_{1}\) and \(X_{3}\). Finally, SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} show a \emph{collider}, where variables \(X_{1}\) and
\(X_{3}\) are concurrent causes of \(X_{2}\). In this configuration,
\(X_{1}\) and \(X_{3}\) are not causally related to each other but both
influence \(X_{2}\) (an \emph{inmorality}). Notably, all building blocks
assume the errors are independent of each other and from all other
variables in the graph, as evidenced by the pairwise relations
\(e_{X1} \perp\!\!\!\perp e_{X2}\), \(e_{X1} \perp\!\!\!\perp e_{X3}\),
and \(e_{X2} \perp\!\!\!\perp e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \perp\!\!\!\perp e_{X} \\
  e_{T} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

@@

\subsection{The probabilistic and causal implications of SCMs and
DAGs}\label{sec-appendix-D}

Beyond their graphical capabilities, SCMs and DAGs can encode the
probabilistic information embedded within a causal structure. They
achieve this encoding by relying on three fundamental assumptions: the
local Markov, the minimality, the causal edges and the modularity
assumptions. The \emph{local Markov assumption} encodes probabilistic
independencies between variables by declaring that nodes in a graph are
independent of all its non-descendants, given its parents
\citep[pp.~20]{Neal_2020}. Meanwhile, the \emph{minimality assumption}
encodes probabilistic dependencies among variables by stating that every
pair of adjacent nodes exhibits a dependency \citep[pp.~21]{Neal_2020}.
Finally, the \emph{causal edges assumption} encodes causal relationships
between variables by declaring that each parent node acts as a direct
cause of its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
illustrates how these assumptions influence the statistical and causal
interpretations of graphs.

@@

The encoding of probabilistic information in SCMs and DAGs facilitates
the efficient representation of the joint distribution of variables
\citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
assumptions underlying the probabilistic encoding is that any conceptual
model described by an SCM and DAG can express the joint distribution of
variables as a product of conditional probability distributions (CPDs)
in the form \(P(child \: | \: parents)\). This property is formally
known as the \emph{Bayesian Network factorization} (BNF)
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this factorization, depicted in Equation~\ref{eq-net_factor},
\(pa(X_{i})\) denotes the set of variables that are the parents of
\(X_{i}\).

\begin{equation}\phantomsection\label{eq-net_factor}{
P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
}\end{equation}

This encoding enables researchers with conceptual (theoretical)
knowledge in the form of an SCM and DAG to predict patterns of
(in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in this data can provide significant insights into the
validity of the proposed conceptual model. Moreover, it allows to
express

>>>>

\newcommand{\dsep}{\:\bot\:}
\newcommand{\ndsep}{\:\not\bot\:}
\newcommand{\cond}{\:|\:}

@@

Empirical research uses evidence from observation and experimentation to
address real-world challenges. In this context, researchers typically
formulate their research questions as \emph{estimands} or \emph{targets
of inference}, i.e., the specific quantities they seek to determine
\citep{Everitt_et_al_2010}. For instance, researchers might be
interested in answering the following question: ``To what extent do
different teaching methods \((T)\) influence students' ability to
produce high-quality written texts \((Y)\)?'' To investigate this,
researchers could randomly assign students to two groups, each exposed
to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
perform pairwise comparisons, generating a dichotomous outcome
\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
ability. In this scenario, the research question can be rephrased as the
estimand, ``\emph{On average}, is there a difference in the ability to
produce high-quality written texts between the two groups of students?''
and this estimand can be mathematically represented by the random
associational quantity in Equation~\ref{eq-group_diff}, where
\(E[\cdot]\) denotes the expected value.

\begin{equation}\phantomsection\label{eq-group_diff}{
E[Y_{i} \:|\:T_{i}=1] - E[Y_{i} \:|\:T_{i}=2]
}\end{equation}

@@

While numerous methods can approximate an estimand, researchers
prioritize estimators with desirable identification properties that
ensure the accuracy of estimates. For instance, the Z-test is an
estimator known for its effectiveness when comparing groups'
proportions, yielding accurate estimates when the underlying assumptions
of the statistic are met \citep{Kanji_2006}. If this is the case, the
Z-test is expressed as a signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i} \:|\:T_{i}=1]\) and \(E[Y_{i} \:|\:T_{i}=2]\), respectively.
The noise, represented by \(\hat{s}_{p}\), is defined as the unpooled
sample variability observed between the two groups.

@@

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key properties, such as
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from any other variable \(X\)
\citep{Morgan_et_al_2014, Neal_2020}.

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendix-A}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes, as in Equation~\ref{eq-ICE}.

\begin{equation}\phantomsection\label{eq-ICE}{
\tau_{i} = Y_{i} \:|\:do(T_{i}=1) - Y_{i} \:|\:do(T_{i}=2)
}\end{equation}

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. In contrast,
conditioning involves restricting the focus to the subset of the
population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).\\

Note that if a student is assigned to intervention \(T_{i}=1\), the
potential outcome under \(T_{i}=2\) becomes a counterfactual, as it is
no longer observed nor observable. To address this challenge, the
structural approach extends the ICE to the \emph{average causal effect}
(ACE, Equation~\ref{eq-ACE}), representing the average difference
between observed potential outcomes and their counterfactual
counterparts.

\begin{equation}\phantomsection\label{eq-ACE}{
\tau = E[\tau_{i}] = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
}\end{equation}

Even though counterfactuals are unobserved and unobservable, researchers
can still identify the ACE from associational estimates by leveraging
the structural approach. The approach identifies the ACE by
statistically conditioning data on a \emph{sufficient adjustment set} of
variables \(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
This \emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. If such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}.

Conditioning on a sufficient adjustment set enables researchers to
estimate the ACE, even when the data comes from an observational study.
This process is feasible because such conditioning ensures that the ACE
estimator satisfies several critical properties, including confounding
elimination \citep{Morgan_et_al_2014}. Naturally, the validity of claims
about the causal effects of \(T\) on \(Y\) now hinges on the assumption
that \(X\) serves as a sufficient adjustment set. However, as
\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
real world from observed data inevitably requires assumptions. This
requirement holds true for both observational and experimental data.

For instance, If researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as a
variable \(X\), such as the socio-economic status of the school,
satisfies the backdoor criterion. Under these circumstances, researchers
firt identify the \emph{conditional average causal effect} (CACE,
Equation~\ref{eq-CACE})

\begin{equation}\phantomsection\label{eq-CACE}{
E[Y_{i} \:|\:T_{i}=t, X]
}\end{equation}

From the CACE, they identify the ACE from associational quantities as in
Equation~\ref{eq-mACE}. Here, \(E_{X}[\cdot]\) represents the marginal
expected value over \(X\) \citep{Morgan_et_al_2014}.

\begin{equation}\phantomsection\label{eq-mACE}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

Notably, the approach extends the ACE identification for a continuous
variable \(T\) as in Equation~\ref{eq-mACE_cont}, ensuring broad
applicability across different causal scenarios
\citep[pp.~45]{Neal_2020}

\begin{equation}\phantomsection\label{eq-mACE_cont}{
\begin{aligned}
  \tau &= E[Y_{i} \:|\:do(T_{i}=t)] \\
  & = d E_{X}\left[ E[Y_{i} \:|\:T_{i}=t, X]\right]/ dt
  \end{aligned}
}\end{equation}

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that facilitate valid causal inference by guiding researchers in
determining which statistical models can identify an estimand (ACE,
CACE, or other), assuming the depicted causal structure is correct
\citep{McElreath_2020}.

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Figure~\ref{fig-IEflow} shows the role of theoretical models in the
inference process.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{3} & := f_{X3}(e_{X3}) \\
  e_{X1} & \:\bot\:e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
  e_{X1} & \:\bot\:e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
  e_{X1} & \:\bot\:e_{X2} \\
  e_{X1} & \:\bot\:e_{X3} \\
  e_{X2} & \:\bot\:e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
  X_{2} & := f_{X2}(e_{X2}) \\
  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
  e_{X1} & \:\bot\:e_{X2} \\
  e_{X1} & \:\bot\:e_{X3} \\
  e_{X2} & \:\bot\:e_{X3}
\end{aligned}
\]

@@

\[
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
  X_{3} & := f_{X3}(e_{X3}) \\
  e_{X1} & \:\bot\:e_{X2} \\
  e_{X1} & \:\bot\:e_{X3} \\
  e_{X2} & \:\bot\:e_{X3}
\end{aligned}
\]

@@

A careful examination of the figures highlights the theoretical
assumptions underlying the observed variables in these building blocks.
SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because there is a \emph{direct path}
connecting. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a
\emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
where variables \(X_{1}\) and \(X_{3}\) are both influenced by
\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, all building blocks assume the errors are
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \:\bot\:e_{X} \\
  e_{T} & \:\bot\:e_{Y} \\
  e_{X} & \:\bot\:e_{Y}
\end{aligned}
\]

@@

\subsection{The probabilistic implications of SCMs and
DAGs}\label{sec-appendix-D}

Beyond their graphical capabilities, SCMs and DAGs can encode the
probabilistic information embedded within a causal structure. They
achieve this encoding by relying on three fundamental assumptions: the
local Markov, the minimality, the causal edges assumptions. The
\emph{local Markov assumption} encodes probabilistic independencies
between variables by declaring that nodes in a graph are independent of
all its non-descendants, given its parents \citep[pp.~20]{Neal_2020}.
Meanwhile, the \emph{minimality assumption} encodes probabilistic
dependencies among variables by stating that every pair of adjacent
nodes exhibits a dependency \citep[pp.~21]{Neal_2020}. Finally, the
\emph{causal edges assumption} encodes causal relationships between
variables by declaring that each parent node acts as a direct cause of
its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
illustrates how these assumptions influence the statistical and causal
interpretations of graphs.

@@

A notable implication of the assumptions underlying the probabilistic
encoding is that any conceptual model described by an SCM and DAG can
represent the joint distribution of variables more efficiently
\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
product of conditional probability distributions (CPDs)
\(P(child \:|\:parents)\). This property is formally known as the
\emph{Bayesian Network factorization} (BNF,
Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this expression, \(pa(X_{i})\) denotes the set of variables that are
the parents of \(X_{i}\).

\begin{equation}\phantomsection\label{eq-net_factor}{
P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \:|\:pa(X_{i}) )
}\end{equation}

This encoding enables researchers with conceptual (theoretical)
knowledge in the form of an SCM and DAG to predict patterns of
(in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in this data can provide significant insights into the
validity of the proposed conceptual model.
